A space-time skew-$t$ model for threshold exceedances

### 2016-01-24 LaTeX/student-competition/skewnormal.tex
cover letter for student paper competition

\subsection{Extremal dependence}\label{s:extdep}
Our interest lies in spatial dependence in the tail of the skew-$t$ process.
One measure of extremal dependence is the $\chi$ statistic \citep{Coles1999}.
For a stationary and isotropic spatial process, the $\chi$ statistic for locations $\bs$ and $\bt$ separated by distance $h = || \bs - \bt ||$ with identical marginal distributions is
\begin{align}
  \chi(h) = \lim_{c \rightarrow c^*} \Pr[Y(\bs) > c | Y(\bt) > c]
\end{align}
where $c^*$ is the upper limit of the support of $Y$.
If $\chi(h) = 0$, then observations are asymptotically independent at distance $h$.
For Gaussian processes, $\chi(h) = 0$ regardless of the distance $h$, so they are not suitable for modeling asymptotically dependent extremes.
Unlike the Gaussian process, the skew-$t$ process is asymptotically dependent (the explicit expression for $\chi(h)$ is given in \ref{a:skewt}).
However, one problem with the spatial skew-$t$ process is that $\lim_{h \rightarrow \infty} \chi(h) > 0$.
This occurs because all observations, both near and far, share the same $z$ and $\sigma$ terms.
Therefore, this long-range dependence feature of the skew-$t$ process is not ideal for spatial analysis of large geographic regions where we expect only local spatial dependence.
We propose a solution to this in Section \ref{s:part}.

@@

The partitioning model remove long-range dependence. Conditional on knots $\bw_1, \ldots, \bw_K$, the $\chi$ statistic for two sites $\bs$ and $\bt$ in partitions $k_s$ and $k_t$ respectively is
\begin{align}
  \chi(h) &= I(k_s = k_t) \chi_{\text{skew-}t}(h) + I(k_s \neq k_t) \chi_{\text{Gaus}}(h) \nonumber \\
         &= I(k_s = k_t) \chi_{\text{skew-}t}(h)
\end{align}
where $I(\cdot)$ is an indicator function, $\chi_{\text{skew-}t}(h)$ is the $\chi$ statistic for a skew-$t$ process given in (\ref{eq:chiskew-t}), $\chi_{\text{Gaus}}(h)$ is the $\chi$ statistic for a Gaussian process, and $h = ||\bs - \bt||$.
Therefore, sites in different subregions are asymptotically independent because $\chi_{\text{Gaus}}(h) = 0$ for all $h$.
Marginally, over the knots, $\chi(h) = \pi(h) \chi_{\text{skew}-t}(h)$, where $\pi(h) = \Pr(k_s = k_t)$ is the probability that two sites separated by distance $h$ are in the same partition.
In \ref{a:proofsamepartition}, we show that $\lim_{h \rightarrow \infty} \pi(h) = 0$, implying $\lim_{h \rightarrow \infty} \chi(h) = 0$.
In Figure \ref{fig:chi}, we give $\chi(h)$ for $K = 1, 3, 5, 10$ partitions for a skew-$t$ distribution with $\alpha = 10$, and 3 degrees of freedom.
In Figure \ref{fig:chi} we estimate $\pi(h)$ through simulation.
% To estimate $\pi(h)$, we generate 500 sites uniformly over the unit-square.
% We then randomly generate 400 different sets of partitions using $K = 3$, $5$, and $10$.
% For each set of knots, we take $\pi(h)$ to be the proportion of sites in the same partition that are separated by distance $h$.
% This plot demonstrates how partitioning helps to reduce extremal dependence as $h$ increases.

@@

\subsection{Results}\label{s:simresults}
We compared the Brier scores for exceeding 4 different thresholds for each dataset.
The thresholds used for the Brier scores are extreme quantiles from the simulated data for $q(0.90)$, $q(0.95)$, $q(0.98)$, and $q(0.99)$.
Figure \ref{fig:simbrierscores} gives the Brier score relative to the Brier score for the Gaussian method calculated as
\begin{align}
  \text{BS}_{\text{rel}} = \frac{\text{BS}_{\text{method}}}{\text{BS}_{\text{Gaussian}}}.
\end{align}
We analyzed the results for the simulation study using a Friedman test at $\alpha = 0.05$ to see if at least one method had a significantly different Brier score.
For Friedman tests that came back with a significant p-value, we conducted a Wilcoxon-Nemenyi-McDonald-Thompson test to see which of the methods had different results.
The full results for the Wilcoxon-Nemenyi-McDonald-Thompson tests are given in \ref{a:pdiffs}.

@@

% \section*{Supplementary Materials}
% Web Appendices A--E are available with this paper at the Biometrics website on Wiley Online Library. \vspace*{-8pt}

>>>>

\subsection{Extremal dependence}\label{s:extdep}
Our interest lies in spatial dependence in the tail of the skew-$t$ process.
One measure of extremal dependence is the $\chi$ statistic \citep{Coles1999}.
For a stationary and isotropic spatial process, the $\chi$ statistic for locations $\bs$ and $\bt$ separated by distance $h = || \bs - \bt ||$ with identical marginal distributions is
\begin{align}
  \chi(h) = \lim_{c \rightarrow c^*} \Pr[Y(\bs) > c | Y(\bt) > c]
\end{align}
where $c^*$ is the upper limit of the support of $Y$.
If $\chi(h) = 0$, then observations are asymptotically independent at distance $h$.
For Gaussian processes, $\chi(h) = 0$ regardless of the distance $h$, so they are not suitable for modeling asymptotically dependent extremes.
Unlike the Gaussian process, the skew-$t$ process is asymptotically dependent \citep{Padoan2011}.
However, one problem with the spatial skew-$t$ process is that $\lim_{h \rightarrow \infty} \chi(h) > 0$.
This occurs because all observations, both near and far, share the same $z$ and $\sigma$ terms.
Therefore, this long-range dependence feature of the skew-$t$ process is not ideal for spatial analysis of large geographic regions where we expect only local spatial dependence.
We propose a solution to this in Section \ref{s:part}.

@@

The partitioning model remove long-range dependence. Conditional on knots $\bw_1, \ldots, \bw_K$, the $\chi$ statistic for two sites $\bs$ and $\bt$ in partitions $k_s$ and $k_t$ respectively is
\begin{align}
  \chi(h) &= I(k_s = k_t) \chi_{\text{skew-}t}(h) + I(k_s \neq k_t) \chi_{\text{Gaus}}(h) \nonumber \\
         &= I(k_s = k_t) \chi_{\text{skew-}t}(h)
\end{align}
where $I(\cdot)$ is an indicator function, $\chi_{\text{skew-}t}(h) > 0$ is the $\chi$ statistic for a skew-$t$ process given in \citet{Padoan2011}, $\chi_{\text{Gaus}}(h)$ is the $\chi$ statistic for a Gaussian process, and $h = ||\bs - \bt||$.
Therefore, sites in different subregions are asymptotically independent because $\chi_{\text{Gaus}}(h) = 0$ for all $h$.
Marginally, over the knots, $\chi(h) = \pi(h) \chi_{\text{skew}-t}(h)$, where $\pi(h) = \Pr(k_s = k_t)$ is the probability that two sites separated by distance $h$ are in the same partition.
It is straightforward to show that $\lim_{h \rightarrow \infty} \pi(h) = 0$, implying $\lim_{h \rightarrow \infty} \chi(h) = 0$.
In Figure \ref{fig:chi}, we give $\chi(h)$ for $K = 1, 3, 5, 10$ partitions for a skew-$t$ distribution with $\alpha = 10$, and 3 degrees of freedom.
In Figure \ref{fig:chi} we estimate $\pi(h)$ through simulation.
% To estimate $\pi(h)$, we generate 500 sites uniformly over the unit-square.
% We then randomly generate 400 different sets of partitions using $K = 3$, $5$, and $10$.
% For each set of knots, we take $\pi(h)$ to be the proportion of sites in the same partition that are separated by distance $h$.
% This plot demonstrates how partitioning helps to reduce extremal dependence as $h$ increases.

@@

\subsection{Results}\label{s:simresults}
We compared the Brier scores for exceeding 4 different thresholds for each dataset.
The thresholds used for the Brier scores are extreme quantiles from the simulated data for $q(0.90)$, $q(0.95)$, $q(0.98)$, and $q(0.99)$.
Figure \ref{fig:simbrierscores} gives the Brier score relative to the Brier score for the Gaussian method calculated as
\begin{align}
  \text{BS}_{\text{rel}} = \frac{\text{BS}_{\text{method}}}{\text{BS}_{\text{Gaussian}}}.
\end{align}
We analyzed the results for the simulation study using a Friedman test at $\alpha = 0.05$ to see if at least one method had a significantly different Brier score.
For Friedman tests that came back with a significant p-value, we conducted a Wilcoxon-Nemenyi-McDonald-Thompson test to see which of the methods had different results.
% The full results for the Wilcoxon-Nemenyi-McDonald-Thompson tests are given in \ref{a:pdiffs}.

@@

% \section*{Supplementary Materials}
% Web Appendices A--E are available with this paper at the Biometrics website on Wiley Online Library. \vspace*{-8pt}
% \renewcommand{\thesection}{Appendix \Alph{section}}
% \setcounter{section}{0}
% \section{Proof that $\lim_{h \rightarrow \infty} \pi(h) = 0$} \label{a:proofsamepartition}
% Consider a homogeneous spatial Poisson process with intensity $\mu$.
% Define $A$ as the circle with center \hbox{$(\bs_1 + \bs_2) / 2$} and radius $h / 2$.
% Then $\bs_1$ and $\bs_2$ are in different partitions almost surely if two or more points are in $A$.
% Let $N(A)$ be the number of points in $A$, and let
% \begin{align*}
%   \mu(A) = \mu |A| = \mu \pi \left(\frac{h}{2}\right)^2 = \lambda h^2.
% \end{align*}
% Then
% \begin{align*}
%   P[N(A) \ge 2] &= 1 - P[N(A) = 0] - P[N(A) = 1]\\
%                 &= 1 - \exp\{-\lambda h^2\} - \lambda h^2 \exp\{-\lambda h^2\} \\
%                 &= 1 - (1 + \lambda h^2) \exp\{-\lambda h^2\}
% \end{align*}
% which goes to one as $h \rightarrow \infty$.

% \section{Simulation study pairwise difference results} \label{a:pdiffs}
% The following tables show the methods that have significantly different Brier scores when using a Wilcoxon-Nemenyi-McDonald-Thompson test.
% In each column, different letters signify that the methods have significantly different Brier scores.
% For example, there is significant evidence to suggest that method 1 and method 4 have different Brier scores at $q(0.90)$, whereas there is not significant evidence to suggest that method 1 and method 2 have different Brier scores at $q(0.90)$.
% In each table group A represents the group with the lowest Brier scores.
% Groups are significant with a familywise error rate of $\alpha = 0.05$.

% \begin{table}[htbp]
%   \centering
%   \caption{Setting 1 -- Gaussian marginal, $K = 1$ knot}
%   \label{tbl:gaussim}
%   \begin{tabular}{|l|ccc|ccc|cccc|ccc|}
%     \cline{2-14}
%     \multicolumn{1}{c}{} & \multicolumn{3}{|c}{$q(0.90)$} & \multicolumn{3}{|c}{$q(0.95)$} & \multicolumn{4}{|c}{$q(0.98)$} & \multicolumn{3}{|c|}{$q(0.99)$} \\
%     \hline
%     Method 1 & A &   &   & A &   &   & A &   &   &   & A & B &   \\
%     \hline
%     Method 2 & A &   &   & A &   &   & A &   &   &   & A &   &   \\
%     \hline
%     Method 3 &   & B &   &   & B &   &   &   & C &   &   & B &   \\
%     \hline
%     Method 4 & A &   &   & A &   &   & A & B &   &   & A & B &   \\
%     \hline
%     Method 5 &   & B &   &   & B &   &   & B & C &   & A & B &   \\
%     \hline
%     Method 6 &   &   & C &   &   & C &   &   &   & D &   &   & C \\
%     \hline
%   \end{tabular}
% \end{table}

% % \begin{table}[htbp]
% %   \centering
% %   \caption{Setting 2: Symmetric-$t$ marginal, $K = 1$ knot}
% %   \label{tbl:t1k1sim}
% %   \begin{tabular}{|l|cc|cccc|cccc|ccc|}
% %     \cline{2-14}
% %     \multicolumn{1}{c}{} & \multicolumn{2}{|c}{$q(0.90)$} & \multicolumn{4}{|c}{$q(0.95)$} & \multicolumn{4}{|c}{$q(0.98)$} & \multicolumn{3}{|c|}{$q(0.99)$} \\
% %     \hline
% %     Method 1 &   & B &   &   &   & D &   &   &   & D &   &   & C \\
% %     \hline
% %     Method 2 & A &   & A &   &   &   & A &   &   &   & A &   &   \\
% %     \hline
% %     Method 3 &   & B &   & B &   &   &   & B & C &   &   & B &   \\
% %     \hline
% %     Method 4 & A &   &   &   & C &   &   & B &   &   &   & B &   \\
% %     \hline
% %     Method 5 &   & B &   &   &   & D &   &   & C & D &   & B & C \\
% %     \hline
% %   \end{tabular}
% % \end{table}

% % \begin{table}[htbp]
% %   \centering
% %   \caption{Setting 3: Symmetric-$t$ marginal, $K = 5$ knots}
% %   \label{tbl:gaussim}
% %   \begin{tabular}{|l|cc|cc|cc|cc|}
% %     \cline{2-9}
% %     \multicolumn{1}{c}{} & \multicolumn{2}{|c}{$q(0.90)$} & \multicolumn{2}{|c}{$q(0.95)$} & \multicolumn{2}{|c}{$q(0.98)$} & \multicolumn{2}{|c|}{$q(0.99)$} \\
% %     \hline
% %     Method 1 &   & B &   & B &   & B &   & B \\
% %     \hline
% %     Method 2 &   & B &   & B &   & B &   & B \\
% %     \hline
% %     Method 3 &   & B &   & B &   & B & A & B \\
% %     \hline
% %     Method 4 & A &   & A &   & A &   & A &   \\
% %     \hline
% %     Method 5 &   & B &   & B & A & B & A & B \\
% %     \hline
% %   \end{tabular}
% % \end{table}

% \begin{table}[htbp]
%   \centering
%   \caption{Setting 2 -- Skew-$t$ marginal, $K = 1$ knot}
%   \label{tbl:st1sim}
%   \begin{tabular}{|l|ccccc|cccc|cccc|ccc|}
%     \cline{2-17}
%     \multicolumn{1}{c}{} & \multicolumn{5}{|c}{$q(0.90)$} & \multicolumn{4}{|c}{$q(0.95)$} & \multicolumn{4}{|c}{$q(0.98)$} & \multicolumn{3}{|c|}{$q(0.99)$} \\
%     \hline
%     Method 1 &   &   & C &   &   &   & B &   &   &   & B & C &   &   & B &   \\
%     \hline
%     Method 2 & A &   &   &   &   & A &   &   &   & A &   &   &   & A &   &   \\
%     \hline
%     Method 3 &   & B & C &   &   & A & B &   &   & A & B &   &   & A & B &   \\
%     \hline
%     Method 4 & A & B &   &   &   &   & B &   &   &   & B &   &   & A &   &   \\
%     \hline
%     Method 5 &   &   &   & D &   &   &   & C &   &   &   & C &   &   & B &   \\
%     \hline
%     Method 6 &   &   &   &   & E &   &   &   & D &   &   &   & D &   &   & C \\
%     \hline
%   \end{tabular}
% \end{table}

% \begin{table}[htbp]
%   \centering
%   \caption{Setting 3 -- Skew-$t$ marginal, $K = 5$ knots}
%   \label{tbl:st5sim}
%   \begin{tabular}{|l|ccc|cccc|ccc|ccc|}
%     \cline{2-14}
%     \multicolumn{1}{c}{} & \multicolumn{3}{|c}{$q(0.90)$} & \multicolumn{4}{|c}{$q(0.95)$} & \multicolumn{3}{|c}{$q(0.98)$} & \multicolumn{3}{|c|}{$q(0.99)$} \\
%     \hline
%     Method 1 &   & B &   &   &   & C &   &   & B &   &   & B &   \\
%     \hline
%     Method 2 &   & B &   &   &   & C &   &   & B &   &   & B &   \\
%     \hline
%     Method 3 & A &   &   &   & B &   &   &   & B &   &   & B &   \\
%     \hline
%     Method 4 & A &   &   & A &   &   &   & A &   &   & A &   &   \\
%     \hline
%     Method 5 & A &   &   & A &   &   &   & A &   &   & A &   &   \\
%     \hline
%     Method 6 &   &   & C &   &   &   & D &   &   & C &   &   & C \\
%     \hline
%   \end{tabular}
% \end{table}

% \begin{table}[htbp]
%   \centering
%   \caption{Setting 4 -- Max-stable}
%   \label{tbl:mssim}
%   \begin{tabular}{|l|cccc|cccc|ccc|ccc|}
%     \cline{2-15}
%     \multicolumn{1}{c}{} & \multicolumn{4}{|c}{$q(0.90)$} & \multicolumn{4}{|c}{$q(0.95)$} & \multicolumn{3}{|c}{$q(0.98)$} & \multicolumn{3}{|c|}{$q(0.99)$} \\
%     \hline
%     Method 1 & A & B &   &   &   & B &   &   &   & B &   &   &   & C \\
%     \hline
%     Method 2 &   & B &   &   &   & B & C &   &   & B &   &   & B & C \\
%     \hline
%     Method 3 &   &   & C & D &   &   & C &   &   & B &   &   & B &   \\
%     \hline
%     Method 4 &   &   &   & D &   &   &   & D &   &   & C &   &   & C \\
%     \hline
%     Method 5 &   &   & C &   &   &   & C &   &   & B &   &   & B & C \\
%     \hline
%     Method 6 & A &   &   &   & A &   &   &   & A &   &   & A &   &   \\
%     \hline
%   \end{tabular}
% \end{table}

% \begin{table}[htbp]
%   \centering
%   \caption{Setting 5 -- Transformation below $T = q(0.80)$}
%   \label{tbl:transsim}
%   \begin{tabular}{|l|cccc|ccc|cccc|cccc|}
%     \cline{2-16}
%     \multicolumn{1}{c}{} & \multicolumn{4}{|c}{$q(0.90)$} & \multicolumn{3}{|c}{$q(0.95)$} & \multicolumn{4}{|c}{$q(0.98)$} & \multicolumn{4}{|c|}{$q(0.99)$} \\
%     \hline
%     Method 1 &   &   & C &   &   & B &   &   &   & C &   &   &   & C &   \\
%     \hline
%     Method 2 &   & B &   &   &   & B &   &   & B &   &   & A & B &   &   \\
%     \hline
%     Method 3 & A &   &   &   & A &   &   & A &   &   &   & A &   &   &   \\
%     \hline
%     Method 4 &   & B & C &   &   & B &   &   & B &   &   &   & B & C &   \\
%     \hline
%     Method 5 &   & B &   &   &   & B &   &   & B & C &   &   &   & C &   \\
%     \hline
%     Method 6 &   &   &   & D &   &   & C &   &   &   & D &   &   &   & D \\
%     \hline
%   \end{tabular}
% \end{table}
